---
title: "Alternative Nulls"
author: "Andrew Park"
date: "5/29/2017"
output: html_document
---

A kind of repository of alternative null ideas (not v standard.)

When using null models to generate standard effect sizes for mean pairwise phylogenetic distance, it is necessary to make choices regarding the sampling frequency of each host species. In the main text, we use the number of host records in GMPD as the sampling frequency. Here, we explore the effect of potential bias associated with commonly-studied hosts appearing in GMPD multiple times per parasite species and per citation. In other words, in some published studies, a given host parasite interaction may result in several GMPD records (stratified by host age, host sex or distinct geographic samples). Here, we only retain unique instances of host species, parasite species and citation so that such stratified cases are recorded only once. 


```{r}
#Charlie's idea...
library(magrittr)
library(dplyr)
library(tibble)
load("get_d4altnull.Rda")
a<-subset(d,select=c("ParasiteCorrectedName","HostCorrectedName","ParType","Citation"))
a<-a[!duplicated(a),]
a<-droplevels(a)
a %<>% as_tibble

t1<-table(d$HostCorrectedName) %>% as_tibble
t2<-table(a$HostCorrectedName) %>% as_tibble
t1t2<-inner_join(t1,t2,by="Var1")
t1t2 %<>% mutate(f1=n.x/sum(n.x))
t1t2 %<>% mutate(f2=n.y/sum(n.y))
t1t2 %>% filter(f1<1) %>% ggplot(.)+geom_point(aes(f1,f2))+geom_abline(intercept=0,slope=1,col="purple")

t1t2 %>% filter(f1<0.02) %>% ggplot(.)+geom_point(aes(f1,f2))+geom_abline(intercept=0,slope=1,col="purple")
```
Fig. S5: Host frequency in the main analysis (f1) and in this supplementary analysis (f2). Disproportionately frequent hosts in the main analysis are relatively less frequent in the supplementary analysis. The purple line shows the 1-to-1 relationship between frequencies.                                                                     




### Test Null

Here I'm hard coding a method that samples hosts according to their abundances to see how close the results are to the ses.mpd approach

```{r}
library(dplyr)

load("get_nri.Rda")
load("get_phydistmini.Rda")
load("get_comm.Rda")

altNull<-nri %>% filter(ntaxa>1) %>% select(ParasiteCorrectedName,ntaxa,mpd.obs,mpd.obs.z) %>% as_tibble

nuller2<-function(df_sr,df_cit,phydist_matrix){
  #runs an alternate null model picking
  #hosts from list of unique H,P,citation
  
  #arguments1: df_sr=data.frame of parasite sp (col1)
  #and host richness (col2)
  #arguments2: df_cit=data.frame of unique
  #H,P,citation records
  #arguments3: phydist_matrix=cophenetic matrix
  #of host phylo dists
  #arguments4: rep number (defaults to 10)
  
  #returns: the original df_sr with standard effect size for mean
  #pairwise phylo dist column appended
  
  df_cit$HostCorrectedName %<>% as.character
  df_sr$av<-(-1) #dummy
  df_sr$sd<-(-1) #dummy

  for (i in 1:dim(df_sr)[1]){
    pdVals<-list()
    for (rep in 1:1000){#how many reps in null model
      n.hosts.target<-df_sr$ntaxa[i]
      n.hosts.sofar<-0
      sample.hosts<-NULL
      while(n.hosts.sofar<n.hosts.target){
        sample.hosts<-c(sample.hosts,sample_n(df_cit,1)$HostCorrectedName)
        sample.hosts <- unique(sample.hosts)
        n.hosts.sofar<-length(sample.hosts)
      }
    pdMtrx<-phydist.mini[sample.hosts,sample.hosts]
    pdVals[[rep]]<-mean(pdMtrx[upper.tri(pdMtrx)])
    }
    av_pdVals<-mean(unlist(pdVals))
    sd_pdVals<-sd(unlist(pdVals))
    df_sr$av[i]<-av_pdVals
    df_sr$sd[i]<-sd_pdVals
  }
  df_sr %<>% mutate(newZ=(mpd.obs-av)/sd)
  
  return(df_sr)
}


```


```{r}
load("get_nri.Rda")
load("get_phydistmini.Rda")

altNull<-nri %>% filter(ntaxa>1) %>% select(ParasiteCorrectedName,ntaxa,mpd.obs,mpd.obs.z) %>% as_tibble

nuller<-function(df_sr,df_cit,phydist_matrix){
  #runs an alternate null model picking
  #hosts from list of unique H,P,citation
  
  #arguments1: df_sr=data.frame of parasite sp (col1)
  #and host richness (col2)
  #arguments2: df_cit=data.frame of unique
  #H,P,citation records
  #arguments3: phydist_matrix=cophenetic matrix
  #of host phylo dists
  #arguments4: rep number (defaults to 10)
  
  #returns: the original df_sr with standard effect size for mean
  #pairwise phylo dist column appended
  
  df_cit$HostCorrectedName %<>% as.character
  df_sr$av<-(-1) #dummy
  df_sr$sd<-(-1) #dummy

  for (i in 1:dim(df_sr)[1]){
    pdVals<-list()
    for (rep in 1:1000){#how many reps in null model
      n.hosts.target<-df_sr$ntaxa[i]
      n.hosts.sofar<-0
      sample.hosts<-NULL
      while(n.hosts.sofar<n.hosts.target){
        sample.hosts<-c(sample.hosts,sample_n(df_cit,1)$HostCorrectedName)
        sample.hosts <- unique(sample.hosts)
        n.hosts.sofar<-length(sample.hosts)
      }
    pdMtrx<-phydist.mini[sample.hosts,sample.hosts]
    pdVals[[rep]]<-mean(pdMtrx[upper.tri(pdMtrx)])
    }
    av_pdVals<-mean(unlist(pdVals))
    sd_pdVals<-sd(unlist(pdVals))
    df_sr$av[i]<-av_pdVals
    df_sr$sd[i]<-sd_pdVals
  }
  df_sr %<>% mutate(newZ=(mpd.obs-av)/sd)
  
  return(df_sr)
}
```

```{r, eval=F}
zzz<-nuller(altNull,a,phydist.mini)
```

```{r}
load("get_altnri.Rda")
```


```{r}
library(ggplot2)
library(gridExtra)

plotS6a <- ggplot(zzz,aes(mpd.obs.z,newZ))+geom_point()+geom_smooth()
plotS6b <- zzz %>% filter(mpd.obs.z>-2) %>% ggplot(.,aes(mpd.obs.z,newZ))+geom_point()+geom_smooth()

grid.arrange(plotS6a,plotS6b, nrow = 2)
```

Fig. S6: The standard effect size measures for mean pairwise phylogenetic distance are highly correlated between the main analysis and supplementary analysis. Specialist parasites of the main analysis are even more highly specific in the supplementary analysis (top panel). For the parasites that were as specific as expected by chance in the original analysis, many of them are considered extremely host specific in the supplementary analysis.

```{r}
zz<-inner_join(zzz,nri,by="ParasiteCorrectedName") %>% as_tibble
zz %<>% mutate(ParType=as.character(ParType))
zz %<>% select(ParasiteCorrectedName,newZ,ParType,close,nonclose,vector,intermediate)
zz %<>% gather(c(close,nonclose,vector,intermediate),key=tmode,value=exhibited) %>% filter(exhibited==1 & ParType %in% c("Arthropod","Bacteria","Helminth","Protozoa","Virus")) %>% select(-exhibited)

ggplot(zz, aes(x=ParType, y=newZ, fill=tmode)) + geom_boxplot(notch=F) + theme(axis.text.x  = element_text(angle=90, vjust=0.5)) + scale_fill_discrete(name="Transmission\nmode",labels=c("Close","Complex","Vector","Envir"))+xlab("Parasite type")+ylab("Standard effect size of mean\npairwise PD between hosts")+geom_hline(yintercept=global.median.z,colour="black")#+ylim(c(-5,1))
```



